Improved Particle Approximations \\ to the Joint Smoothing Distribution \\ Using Markov Chain Monte Carlo

### 2012-02-15 Paper/draft5_MCMCsmoother.tex
Rewriting draft.

\IEEEPARstart{T}{he} objective of sequential Bayesian inference is the estimation of an unknown, time-varying quanitity from incomplete or inaccurate observations. This is commonly achieved through the use of probabilistic models for the evolution of the latent state and the measurement process.

@@

Here, the random variable $x_k$ denotes the value of the latent state at the $k^{th}$ instant, and $y_k$ the value of the observation. A set of random variables will be written as $x_{1:k} = \{x_1, x_2, \dots, x_k \}$. The state evolution is assumed to be Markovian, i.e. $x_k$ depends only on the previous value, $x_{k-1}$.

@@

In the case where the state and observation processes of equations~\ref{eq:state_proc}~and~\ref{eq:observ_proc} are linear and Gaussian, the filtering problem may be solved analytically using the Kalman filter \cite{Kalman1960}. For nonlinear or non-Gaussian models, tractable closed-form solutions are rare. In such cases, numerical approximations are often employed, including the particle filter algorithm, first introduced by \cite{Gordon1993}. (See \cite{Cappe2007,Doucet2009} for a thorough introduction.)

@@

Here, $\delta_{a}(x)$ indicates a discrete probability mass at the point $x = a$. $\hat{p}$ means an approximation to $p$.

A similar story applies to the problem of smoothing. In the linear Gaussian case, the entire state sequence, $x_{1:K}$, may be estimated in closed-form using the a Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}. This works by modifying the ordinary Kalman filter results in a backward processing pass through the observations. For nonlinear or non-Gaussian models, it is possible again to resort to numerical methods. Now, the particles of the approximation are drawn from the joint smoothing distribution.

@@

Each particle is an entire realisation of the state process, with $K$ values corresponding to the state at each time step. Methods for generating these samples have been described in \cite{Kitagawa1996,Godsill2004,Briers2010}. In this paper new methods are presented for drawing samples from the joint smoothing distribution, using Markov chain Monte Carlo (MCMC). These novel procedures have computational advantages and greater flexibility over previous methods.

@@

The particle filter is a recursive numerical algorithm for estimation of the filtering distribution. At the $k^{th}$ processing step, a particle estimate of the joint distribution over $x_{k-1}$ and $x_k$ is made by drawing samples from a proposal distribution.

\begin{IEEEeqnarray}{rCl}
\{ x_{k-1}^{(i)}, x_k^{(i)} \} & \sim & q(x_{x}|x_{k-1}) q(x_{k-1})
\end{IEEEeqnarray}

The $k-1$ state proposal distribution, $q(x_{k-1})$, is constructed from the particles of the filtering distribution from the previous time step. The choice of weights and sampling procedure here determines what from of ``resampling'' is to be used. Here we use arbitrary weights for generality.

@@

Note that the choice $v_{k-1}^{(i)} = w_{k-1}^{(i)}$ corresponds to ordinary resampling, and other choices of $v_{k-1}^{(i)}$ represent auxiliary sampling schemes \cite{Pitt1999}. For the least computational expense, the proposal distribution with weights $v_{k-1}^{(i)} = 1/N_F$ (where $N_F$ is the number of filter particles) can be ``sampled'' by simply keeping the set of particles generated in the last step, with no resampling. (See \cite{Cappe2007,Doucet2009} for further discussion of resampling.)

Next, the particles are assigned an importance weight according to the ratio of the targeted joint distribution and the proposal.

@@

Normalisation is enforced by scaling the weights so that they sum to 0. Finally, $x_{k-1}$ is marginalised from the distribution by simply discarding the values from each particles.

@@

The weakness of this simple, filter-smoother (FS) is particle degeneracy. Every time the particles are resampled, those with low weights do not appear in the approximation at the next step, while those with high weights appear multiple times. The result is that many, or even all, particles will have the same values at early time steps. This effect is illustrated in figure~\ref{} in section~\ref{}. If the objective is only to produce a single point estimate then this might be sufficient, but to characterise the complete smoothing distribution the approximation must be rejeuvenated.

@@

Particles are generated by sampling backwards in time from the factors of this expansion. At the $k^{th}$ step, a sample $\tilde{x}_{k+1:K} \sim p(x_{k+1:K}|y_{1:K})$ will already have been drawn. By sampling $x_k \sim p(x_k|\tilde{x}_{k+1}, y_{1:K})$ and appending $x_k$ to $\tilde{x}_{k+1:K}$, the state sequence is sequentially extended back in time. The recursion is initialised with $x_K \sim p(x_K|y_{1:K})$, i.e. a sample from the final filtering distribution.

The backwards conditional distributions may be expanded with Bayes rule.

@@

A method for drawing samples from the joint smoothing distribution, $p(x_{1:K}|y_{1:K})$, is included in \cite{Briers2010}. The principal topic of this paper is a two-filter smoother for estimation of the marginal smoothing distributions, $p(x_{k}|y_{1:K})$, by combination of the results from one forward and one backward particle filter. Particles from the joint distribution may then be generated by resampling from either the forward or backward filters.% in the manner of algorithm~\ref{alg:DBRS}.

@@

In the algorithms described so far, no new states are sampled during the smoothing stage; there is only a resampling of the filtering distributions. This limits the support of the smoother to the states that appear in the filtering particles. In \cite{Fearnhead2010}, an algorithm for estimation of the marginal smoothing distributions is described in which new states are sampled from a proposal, allowing full support. In addition, this algorithm achieves $\mathcal{O}( N_S \times K)$ complexity, linear in the number of particles!

@@

The bottleneck of the backward-resampling smoother is the calculation of the sampling weights (equation~\ref{eq:DBRS_weights}). For each smoother particle and for each time step, a weight must be calculated for each of the $N_F$ filter particles, in order to draw a sample from the conditional distribution $\hat{p}(x_k|\tilde{x}_{k+1}, y_{1:K})$. The innovation here is to use Markov chain Monte Carlo (MCMC) to produce these samples instead of sampling directly.

MCMC algorithms are a class of numerical procedures which use a Markov chain to draw dependent samples from a target probability distribution. A thorough introduction can be found in \cite{Gilks1996}. The basic Metropolis-Hastings (MH) \cite{Metropolis1953} sampler uses a series of steps in which a new state is drawn from a proposal distribution and accepted with a probability determined by the ratios of the target and proposal probabilities.

For the direct backward sampling smoother, the objective at each time step was to draw a sample from $p(x_{k}|\tilde{x}_{k+1}, y_{1:K})$. To simplify the notation and initialisation of chains, the MCMC version will target $p(x_{1:k}|\tilde{x}_{k+1}, y_{1:K})$. Theoretically this means using the particles of the filter-smoother, $\{x_{1:k}^{(i)}\}$, rather than just the filter, $\{x_k^{(j)}\}$, which would require $\mathcal{O}(K^2)$ memory. However, as discussed later, this is not required in practice.

As before, the algorithm begins by first running a particle filter-smoother and then resampling states backwards through time to produce particles from the joint smoothing distribution, using the factorisation of equation~\ref{eq:smoothing_factorisation}. At the $k^{th}$ time step, a Markov chain is used to sample from $p(x_{1:k}|\tilde{x}_{k+1}, y_{1:K})$. The chain is initialised with the first $k$ states from the previous processing step. As this is itself a sample from the target distribution, no burn-in period is required.

Following an equivalent derivation to equation~\ref{eq:backward_conditional_filter}, the target distribution may be approximated by a particle distribution.

@@

A valid proposal distribution may be constructed using the particles of the filter-smoother approximation with any arbitrary weights.

@@

\begin{IEEEeqnarray}{rCl}
\alpha &=& \frac{ \hat{p}(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) }{ \hat{p}(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) } \nonumber \\
 &=& \frac{ w_k^{*} \tilde{v}_k^{(m)} }{  w_k^{(m)} \tilde{v}_k^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) }  .
\end{IEEEeqnarray}

@@

%\begin{algorithm}
\begin{algorithmic}
 	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$.
	\FOR{$i = 1 \dots N_S$}
		\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
		\FOR{$k = K-1 \dots 1$}
			\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
			\FOR{$m = 1 \dots M$}
				\STATE Propose a new state, $x_{1:k}^{(i)*} \sim \sum_j \tilde{v}_k^{(j)} \delta_{x_{1:k}^{(j)}}(x_{1:k})$.
				\STATE With probability $\alpha = \frac{ w_k^{*} \tilde{v}_k^{(m)} }{  w_k^{(m)} \tilde{v}_k^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) }$, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
			\ENDFOR
			\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{k}^{(i)(M)}$
		\ENDFOR
	\ENDFOR
\end{algorithmic}
%\caption{MCMC backward-resampling smoother algorithm}
%\label{alg:MCMC-BRS}
%\end{algorithm}

For a practical implementation, it is not necessary to store the complete particles of the filter-smoother approximations, $\hat{p}(x_{1:k}|y_{1:k})$, which would require memory space which scaled quadratically as $K$ increased. As the smoother progresses backwards in time, the earlier states in the sequence are repeatedly overwritten, and so it is not necessary ever to have known them. Thus, it is only necessary to store the last two states in each filter-smoother particle (i.e. $\{x_{k-1}^{(j)}\}$ and $\{x_{k}^{(j)}\}$).

@@

With the direct backward-resampling smoother, the only parameter to be chosen is $N_S$, the number of smoother particles. With the MCMC variant, we also have the freedom to set $M$, the number of MH steps in each Markov chain. This controls a trade-off between speed and particle diversity. If the chains are short, and the acceptance probabilities low, then this scheme will simply reproduce particles from the filter-smoother. However, when the weights of the conditional distribution approximation (equation~\ref{eq:MCMC-BRS_weights}) are similar, then an independent sample may be produced with only a very short chain.

\section{Improving the smoother support} \label{sec:new_state_smoother}

In \cite{Fearnhead2010}, the authors described an algorithm for estimating the marginal smoothing distributions. For this method, states were proposed afresh rather than simply resampling from the filtering approximations. This yielded improvements in accuracy and particle diversity over other marginal smoothing techniques, and also the advantage of $\mathcal{O}(N_S \times K)$ complexity. Here we adapt the method of \cite{Fearnhead2010} for generating particles from the joint smoothing distribution. Again, an MCMC procedure is employed.

The method described in \cite{Fearnhead2010} uses a forward and a backward particle filter. To develop a marginal smoothing approximation at time step $k$, particles are resampled from the forward filter at step $k-1$, and from the backward filter and step $k+1$. A new state for step $k$ is then sampled from a continuous proposal distribution conditional on the previous and next states, and an importance weight is assigned. It is undesirable to use importance sampling for generating particles from the joint sampling distribution, as the required resampling is liable to lead to a loss of particle diversity similar to that suffered by the filter-smoother.

>>>>

\IEEEPARstart{T}{he} objective of sequential Bayesian inference is the estimation of an unknown, time-varying quanitity from incomplete or inaccurate observations. This is commonly achieved through the use of discrete-time probabilistic models for the evolution of the latent state and the measurement process, defined by transition and observation densities.

@@

Here, the random variable $x_k$ denotes the value of the latent state at the $k$th instant, and $y_k$ the value of the observation. A set of random variables will be written as $x_{1:k} = \{x_1, x_2, \dots, x_k \}$. The state evolution is assumed to be Markovian, i.e. $x_k$ depends only on the previous value, $x_{k-1}$.

@@

In the case where the state and observation processes of (\ref{eq:state_proc}) and~(\ref{eq:observ_proc}) are linear and Gaussian, the filtering problem may be solved analytically using the Kalman filter \cite{Kalman1960}. For nonlinear or non-Gaussian models, tractable closed-form solutions are rare. Instead, numerical approximations are often employed, including the particle filter algorithm, first introduced by \cite{Gordon1993}. (See \cite{Cappe2007,Doucet2009} for a thorough introduction.)

@@

Here, $\delta_{a}(x)$ indicates a discrete probability mass at the point $x = a$.% $\hat{p}$ means an approximation to $p$.

A similar story applies to the problem of smoothing. In the linear Gaussian case, the entire state sequence, $x_{1:K}$, may be estimated in closed-form using the a Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}. This works by modifying the ordinary Kalman filter results in a backward processing pass through the state sequence. For nonlinear or non-Gaussian models, it is possible again to resort to numerical methods. Now, particles are used to approximate the joint smoothing distribution.

@@

Each particle is an entire realisation of the state process, with $K$ values corresponding to the state at each time step. Methods for generating these samples have been described in \cite{Kitagawa1996,Godsill2004,Briers2010}. In this paper new algorithms are presented for drawing samples from the joint smoothing distribution, using Markov chain Monte Carlo (MCMC). These novel procedures have computational advantages and greater flexibility over previous methods.

@@

The particle filter is a recursive numerical algorithm for estimation of the filtering distribution. At the $k$th processing step, a particle estimate of the joint density over $x_{k-1}$ and $x_k$ is made using importance sampling (IS). A particle is first sampled from the factorised proposal density,

\begin{IEEEeqnarray}{rCl}
\{ x_{k-1}^{(i)}, x_k^{(i)} \} & \sim & q(x_{x}|x_{k-1}) q(x_{k-1}).
\end{IEEEeqnarray}

The proposal density for $x_k$, $q(x_{x}|x_{k-1})$, is often called the ``importance density''. Common choices include the model transition density (\ref{eq:state_proc}), and the ``optimal importance distribution'' \cite{Doucet2000a}. The proposal density for $x_{k-1}$, $q(x_{k-1})$, is constructed from the particles of the filtering approximation from the previous time step. The choice of weights and sampling procedure for $x_{k-1}$ determines what from of ``resampling'' is to be used. Here we use arbitrary weights for generality.

@@

the choice $v_{k-1}^{(i)} = w_{k-1}^{(i)}$ corresponds to ordinary resampling, and other choices of $v_{k-1}^{(i)}$ represent auxiliary sampling schemes \cite{Pitt1999}. For the least computational expense, the proposal density with weights $v_{k-1}^{(i)} = 1/N_F$ (where $N_F$ is the number of filter particles) can be ``sampled'' by simply keeping the set of particles generated in the last step, with no resampling. (See \cite{Douc2005} for further discussion of resampling.)

Next, the particles are assigned an importance weight according to the ratio of the target and the proposal densities.

@@

Normalisation is enforced by scaling the weights so that they sum to 0. Finally, $x_{k-1}$ is marginalised from the distribution by simply discarding the values from each particle.

@@

The weakness of this simple, filter-smoother (FS) is degeneracy of the particles in the path-space. Every time the particles are resampled, those with low weights do not appear in the approximation at the next step, while those with high weights appear multiple times. The result is that many, or even all, particles will share a common ancestor, before which they all have the same set of state values. If the objective is only to produce a single point estimate then this might be sufficient, but to characterise the complete smoothing distribution, the path-space diversity of the approximation must be improved.

@@

Particles are generated by sampling backwards in time from the factors of this expansion. At the $k$th step, a sample $\tilde{x}_{k+1:K} \sim p(x_{k+1:K}|y_{1:K})$ will already have been drawn. By sampling $x_k \sim p(x_k|\tilde{x}_{k+1}, y_{1:K})$ and appending $x_k$ to $\tilde{x}_{k+1:K}$, the state sequence is sequentially extended back in time. The recursion is initialised with $x_K \sim p(x_K|y_{1:K})$, i.e. a sample from the final filtering approximation.

The backwards conditional density may be expanded with Bayes rule.

@@

A method for drawing samples from the joint smoothing distribution, $p(x_{1:K}|y_{1:K})$, is included in \cite{Briers2010}. The principal topic of this paper is a two-filter smoother for estimation of the marginal smoothing distributions, $p(x_{k}|y_{1:K})$, by combination of the results from one forward and one backward particle filter. Particles from the joint distribution may then be generated by resampling from either the forward or backward filters (or some combination) in the manner of the backward-resampling smoother outlined above.%algorithm~\ref{alg:DBRS}.

@@

In the algorithms described so far, no new states are sampled during the smoothing stage; there is only a resampling of the filtering distributions. This limits the support of the smoother to the states that appear in the filtering particles. In \cite{Fearnhead2010}, an IS-based algorithm for estimation of the marginal smoothing distributions is described in which new states are sampled from a proposal, allowing full support. In addition, this algorithm achieves $\mathcal{O}( N_S \times K)$ complexity, linear in the number of particles.

@@

The bottleneck of the backward-resampling smoother is the calculation of the sampling weights, specified by (\ref{eq:DBRS_weights}). For each smoother particle and for each time step, a weight must be calculated for each of the $N_F$ filter particles, in order to draw a sample from the conditional distribution $\hat{p}(x_k|\tilde{x}_{k+1}, y_{1:K})$. The innovation here is to use Markov chain Monte Carlo (MCMC) to produce these samples instead of sampling directly.

MCMC algorithms are a class of numerical procedures which use a Markov chain to draw dependent samples from a target probability distribution. A thorough introduction can be found in \cite{Gilks1996}. The basic Metropolis-Hastings (MH) \cite{Hastings1970} sampler uses a series of steps in which a new state is drawn from a proposal distribution and accepted with a probability determined by the ratios of the target and proposal densities.

For the direct backward-resampling smoother, the objective at each time step was to draw a sample from $p(x_{k}|\tilde{x}_{k+1}, y_{1:K})$. To simplify the notation and initialisation of chains, the MCMC version will target $p(x_{1:k}|\tilde{x}_{k+1}, y_{1:K})$. Theoretically this means using the particles of the filter-smoother, $\{x_{1:k}^{(i)}\}$, rather than just the filter, $\{x_k^{(j)}\}$, which would require $\mathcal{O}(K^2)$ memory. However, as discussed later, this is not required in practice.

As before, the algorithm begins by first running a particle filter-smoother and then resampling states backwards through time to produce particles from the joint smoothing distribution, using the factorisation of (\ref{eq:smoothing_factorisation}). At the $k$th time step, a Markov chain is used to sample from $p(x_{1:k}|\tilde{x}_{k+1}, y_{1:K})$. The chain is initialised with the output from the previous processing step. As this is itself a sample from the target distribution, no burn-in period is required.

Following an equivalent derivation to that for (\ref{eq:backward_conditional_filter}), the target distribution may be approximated by a particle distribution.

@@

A valid proposal may be constructed using the particles of the filter-smoother approximation with any arbitrary weights.

@@

\begin{IEEEeqnarray}{rCl}
\alpha &=& \min \Bigg[ 1, \frac{ \hat{p}(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) }{ \hat{p}(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) } \Bigg] \nonumber \\
       &=& \min \Bigg[1, \frac{ w_k^{*} \tilde{v}_k^{(m)} }{  w_k^{(m)} \tilde{v}_k^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) }  \Bigg]  . \label{eq:MCMC-BRS_ap}
\end{IEEEeqnarray}

@@

%\begin{algorithm}
\begin{algorithmic}
 	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$.
	\FOR{$i = 1 \dots N_S$}
		\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
		\FOR{$k = K-1 \dots 1$}
			\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
			\FOR{$m = 1 \dots M$}
				\STATE Propose a new state, $x_{1:k}^{(i)*} \sim \sum_j \tilde{v}_k^{(j)} \delta_{x_{1:k}^{(j)}}(x_{1:k})$.
				\STATE With probability $\alpha$ specified by (\ref{eq:MCMC-BRS_ap}),\\ $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
			\ENDFOR
			\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{1:k}^{(i)(M)}$
		\ENDFOR
	\ENDFOR
\end{algorithmic}
%\caption{MCMC backward-resampling smoother algorithm}
%\label{alg:MCMC-BRS}
%\end{algorithm}

For a practical implementation, it is not necessary to store the complete particles of the filter-smoother approximations, $\hat{p}(x_{1:k}|y_{1:k})$, which would require memory space which scaled quadratically with $K$. As the smoother progresses backwards in time, the earlier states in the sequence are repeatedly overwritten, and so it is not necessary ever to have known them. Thus, it is only necessary to store the last two states in each filter-smoother particle (i.e. $\{x_{k-1}^{(j)}\}$ and $\{x_{k}^{(j)}\}$).

@@

With the direct backward-resampling smoother, the only parameter to be chosen is $N_S$, the number of smoother particles. With the MCMC variant, we also have the freedom to set $M$, the number of MH steps in each Markov chain. This controls a trade-off between speed and accuracy/particle diversity. If the chains are short, and the acceptance probabilities low, then this scheme will simply reproduce particles from the filter-smoother. However, when the weights of the conditional distribution approximation (\ref{eq:MCMC-BRS_weights}) are similar, then an independent sample may be produced with only a very short chain.

\section{An MCMC particle smoother with improved support} \label{sec:new_state_smoother}

In \cite{Fearnhead2010}, the authors described an algorithm for estimating the marginal smoothing distributions. For this method, states were proposed afresh rather than simply resampling from the filtering approximations. This yielded improvements in accuracy and particle diversity over other marginal smoothing techniques, and also the advantage of $\mathcal{O}(N_S \times K)$ complexity. Here we adapt this method for generating particles from the joint smoothing distribution. Again, an MCMC procedure is employed.

%The method described in \cite{Fearnhead2010} uses a forward and a backward particle filter. To develop a marginal smoothing approximation at time step $k$, particles are resampled from the forward filter at step $k-1$, and from the backward filter and step $k+1$. A new state for step $k$ is then sampled from a continuous proposal distribution conditional on the previous and next states, and an importance weight is assigned. It is undesirable to use importance sampling for generating particles from the joint sampling distribution, as the required resampling is liable to lead to a loss of particle diversity similar to that suffered by the filter-smoother.

The smoother described in \cite{Fearnhead2010} uses importance sampling to approximate the marginal smoothing distributions. This method could be extended to estimation of the joint smoothing distribution by applying it sequentially. However, the resampling required to limit weight degeneracy would lead to a loss of path-space diversity similar to that suffered by the filter-smoother.
